Liveness-Based Garbage Collection for Lazy Languages

### 2015-06-05 paper/paper.tex
Updated print stack subsection

In this algorithm we consult the liveness automata only if it is a cons cell, for closures we 
assume everything under it to be live and copy everything that is reachable. Although this might 
lead to copying more cells it is faster as there are no liveness checks. 
\comment{Once experimental results are obtained compare the two algorithms here}
Since we mix both liveness based GC and reachability the garbage collector has to be more careful 
and ensure references marked non-live are never copied using reachability. We discuss this issue 
and its solution in sub-section[]\cred{hanging like a sore thumb}.

@@

\subsubsection{Issues with mixing LGC and RGC}
A consequence of  copying everything under a closure  is the fact that
we can end up trying to copy things which are not live and corrupt the
heap. This  can happen if at some  $GC_k$ a cons cell  is copied using
liveness. At  a later stage this  cons cell becomes part  of a closure
and  hence during $GC_{k+n}$  the algorithm  tries to  copy everything
under the  cons cell  thus copying heap  locations which might  not be
valid.  To avoid  this behavior we either have to choose  to do a full
liveness based  GC for closures or  we should ensure  that any non-live
references are  not carried across  GC's. We take the  latter approach
and ensure that  non-live references are set to  null. After every GC,
the live part of the buffer  is scanned and any reference which is not
live is  set to null.  The correctness of  this follows from  the fact
that any reference which is decalred  not live in a $GC_k$ will not be
derefernced during later execution.
% A reference copied using LGC could become part of a closure in a
later GC
% and copy wrong locations. Avoiding it by setting non-live references
to null after every GC.
% Proof for the correctness.
\subsection{Garbage collection for references on print stack}
Another  difference that is  present in  GC of  lazy language  is the
presence  of  a print  stack.  In  lazy  languages the  evaluation  of
closures is  driven by the print  function\comment{Refer to SPJ's
book}.  Hence, during  a GC there
will be heap  references that will be present on  the stack which also
have  to  copied/collected.   One  solution  is  to  assume  that  all
references  present  on  the  print  stack are  fully  live  and  copy
everything pointed  by them  akin to reachability.   This is  safe but
copies heap  cells which  have already been  printed. Since  the print
function is  not part of  the executing program its  liveness automata
are not  considered. Therefore we  simulate the liveness of  the print
function by using a flag in  the heap cell. The flag ensures that only
the required parts  of a cons cell are copied.\cred{Very sketchy: Make
  another attempt to write}

>>>>

In this algorithm we consult the liveness automata only if it is a cons cell, for closures we 
assume everything under it to be live and copy everything that is reachable. Although this might 
lead to copying more cells it is faster as there are no liveness checks. 
\comment{Once experimental results are obtained compare the two algorithms here}
Mixing liveness based GC and reachability has its own set of pros and cons which 
we discuss further.\cred{hanging like a sore thumb}.

@@

\subsubsection{Issues with mixing LGC and RGC}
A consequence of  copying everything under a closure  is the fact that
we can end up trying to copy things which are not live and corrupt the
heap. This  can happen if at some  $GC_k$ a cons cell  is copied using
liveness. At  a later stage this  cons cell becomes part  of a closure
and  hence during $GC_{k+n}$  the algorithm  tries to  copy everything
under the  cons cell  thus copying heap  locations which might  not be
valid.  To avoid  this behavior we either have to choose  to do a full
liveness based  GC for closures or  we should ensure  that any non-live
references are  not carried across  GC's. We take the  latter approach
and ensure that  non-live references are set to  null. After every GC,
the live part of the buffer  is scanned and any reference which is not
live is  set to null.  The correctness of  this follows from  the fact
that any reference which is decalred  not live in a $GC_k$ will not be
derefernced during later execution.
% A reference copied using LGC could become part of a closure in a
later GC
% and copy wrong locations. Avoiding it by setting non-live references
to null after every GC.
% Proof for the correctness.
\subsection{Garbage collection for references on print stack}
Another  difference that is  present in  GC of  lazy language  is the
presence  of  a print  stack.  In  lazy  languages the  evaluation  of
closures is  driven by the print  function\comment{Refer to SPJ's
book}.  
Printing of atomic values is simple and does not trigger a garbage collection.
In case of \CONS\ cells, the \CAR\ part is first printed and then the \CDR\ part.
In a lazy language, both the \CAR\ and the \CDR\ parts could be closures, thus 
requiring evaluation. A garbage collection might be triggered while printing 
the \CAR\ part or the \CDR\ part. Therefore, whenever a garbage collection is 
triggered we have to consider any references that might be on the print stack 
and copy them. Liveness automata does not exist for the print function as it 
is not part of the executing program. Since the print function makes everything 
reachable from a cell live we can copy all cells reachable from the reference.

Consider the case where a \CONS\ cells is being printed and a garbage collection 
is triggered when printing the \CDR\ part. In this case, since the refernce to the 
\CONS\ cells is still on the print stack both the \CAR\ and the \CDR\ parts are 
reachable but we know that the \CAR\ part has already been printed and can be 
safely garbage collected. A liveness analysis of the print function would have 
given the same information. In our implementation we mimic the behavior of 
the liveness analysis by using a flag in every \CONS\ cell which says whether 
the \CAR\ part has been printed or not. For any \CONS\ cell if the \CAR\ part 
has not been printed, a reference to the \CAR\ part will also be on the stack but 
the \CDR\ part has to be explicitly copied.  
In case the \CAR\ part has been printed, then a reference to the \CDR\ part will 
be on the stack and since the \CAR\ part is not required anymore we do not 
copy anything. This optimization subsumes tail call optimization. Again, 
experimental results show that this optimization minimizes a lot of copying. 
\cred{Very sketchy: Make another attempt to write}
